---
title: "Group Project_4830"
output: html_document
date: "2024-11-11"
---

```{r}
#Load libraries
library(tidyverse)
library(kableExtra)
library(rsample)
library(recipes)
library(parsnip)
library(yardstick)
library(viridisLite)
library(GGally)
```


```{r}
Heart<-read.csv("processed.cleveland.data")

View(Heart)
```

```{r}
names <- c("Age",
           "Sex",
           "Chest_Pain_Type",
           "Resting_Blood_Pressure",
           "Serum_Cholesterol",
           "Fasting_Blood_Sugar",
           "Resting_ECG",
           "Max_Heart_Rate_Achieved",
           "Exercise_Induced_Angina",
           "ST_Depression_Exercise",
           "Peak_Exercise_ST_Segment",
           "Num_Major_Vessels_Flouro",
           "Thalassemia",
           "Diagnosis_Heart_Disease")
```

```{r}
head(Heart)
```

```{r}
colnames(Heart) <- names
View(Heart)
```

```{r}
nrow(Heart)
```

```{r}
missing_values <- sapply(Heart, function(x) sum(is.na(x)))

print(missing_values)
```

```{r}
Heart[Heart == '?'] <- NA
View(Heart)
```


```{r}
missing_values <- sapply(Heart, function(x) sum(is.na(x)))

print(missing_values)
```


```{r}
Heart <- na.omit(Heart)
nrow(Heart)
```

```{r}
#Determine the number of values in each level of dependent variable
Heart %>% 
  drop_na() %>%
  group_by(Diagnosis_Heart_Disease) %>%
  count() %>% 
  ungroup() %>%
  kable(align = rep("c", 2)) %>% kable_styling("full_width" = F)
```

Since any value above 0 in ‘Diagnosis_Heart_Disease’ (column 14) indicates the presence of heart disease, we can lump all levels > 0 together so the classification predictions are binary – Yes or No (1 or 0).
```{r}
Heart$Diagnosis_Heart_Disease <- ifelse(Heart$Diagnosis_Heart_Disease > 0, 1, 0)

head(Heart)
```
```{r}
unique(Heart$Num_Major_Vessels_Flouro)
```


Gender distribution of the variables.

```{r}
categorical_data <- Heart %>% 
  mutate(Sex = if_else(Sex == 1, "MALE", "FEMALE"),
         Exercise_Induced_Angina = if_else(Exercise_Induced_Angina == 1, "YES", "NO"),
         Chest_Pain_Type = if_else(Chest_Pain_Type == 1, "TYPICAL ANGINA",
                                    if_else(Chest_Pain_Type == 2, "ATYPICAL ANGINA",
                                            if_else(Chest_Pain_Type == 3, "NON-ANGINAL PAIN", "ASYMPTOMATIC"))),
         Resting_ECG = if_else(Resting_ECG == 0, "NORMAL",
                               if_else(Resting_ECG == 1, "ABNORMALITY", "PROBABLE OR DEFINITE")),
         Num_Major_Vessels_Flouro = as.factor(Num_Major_Vessels_Flouro),
         Thalassemia = as.factor(Thalassemia),
         Diagnosis_Heart_Disease = if_else(Diagnosis_Heart_Disease == 1, "YES", "NO"),
         Age_Group = case_when(
           Age < 30 ~ "Under 30",
           Age >= 30 & Age < 50 ~ "30-49",
           Age >= 50 & Age < 70 ~ "50-69",
           Age >= 70 ~ "70 and above"
         )
  ) %>%
  mutate_if(is.character, as.factor) %>%
  dplyr::select(Diagnosis_Heart_Disease, Sex, Exercise_Induced_Angina, Chest_Pain_Type, Resting_ECG, Num_Major_Vessels_Flouro, Thalassemia, Age_Group)
```

```{r}
summary(categorical_data)
```

```{r}
str(Heart)
```


```{r}
for (pair in colnames(categorical_data)) {
  
  tbl <- with(categorical_data, table(categorical_data[['Sex']], categorical_data[[pair]]))
  
  barplot(tbl, beside = TRUE, 
          col = c("skyblue", "salmon"), 
          main = paste("Comparison of", pair, "by", 'Sex'), 
          xlab = 'Sex', 
          ylab = "Count")
  
  legend("topright", legend = c('Female','Male'), fill = c("skyblue", "salmon"))
  }
```

In the dataset, the number of males are more than female around 200 patients are male and around 100 is only female.

```{r}
for (pair in colnames(categorical_data)) {
  
  tbl <- with(categorical_data, table(categorical_data[['Diagnosis_Heart_Disease']], categorical_data[[pair]]))
  
  barplot(tbl, beside = TRUE, 
          col = c("skyblue", "salmon"), 
          main = paste("Comparison of", pair, "by", 'Sex'), 
          xlab = 'Sex', 
          ylab = "Count", cex.names = 0.8)
  
  legend("topright", legend = c('0','1'), fill = c("skyblue", "salmon"))
  }
```


Density plots of the numerical variables.

```{r}
# Loop through each numerical variable for density plots
numerical_var = c('ST_Depression_Exercise','Age','Max_Heart_Rate_Achieved','Serum_Cholesterol','Resting_Blood_Pressure')
for (var in numerical_var) {
  p <- ggplot(Heart, aes_string(x = var, color = "factor(Diagnosis_Heart_Disease)")) +
    geom_density(alpha = 0.6, size = 1) +
    labs(title = paste("Density of", var, "by Heart Disease Diagnosis"),
         x = var,
         y = "Density",
         color = "Heart Disease") +
    theme_minimal()
    print(p)
}
```

Boxplot

```{r}
numerical_data = Heart[numerical_var]
boxplot(numerical_data, cex.axis=0.5)
```



```{r}
for (var in numerical_var) {
  p <- ggplot(Heart, aes_string(x = as.factor(Heart$Sex), y = var, fill = as.factor(Heart$Sex))) +
    geom_boxplot(outlier.color = "red", outlier.shape = 16) +
    labs(title = paste("Boxplot of", var, "by Heart Disease Diagnosis"),
         x = "Gender",
         y = var) +
    theme_minimal()
    print(p)
}

```

```{r}
for (var in numerical_var) {
  p <- ggplot(Heart, aes_string(x = as.factor(Heart$Diagnosis_Heart_Disease), y = var, fill = as.factor(Heart$Diagnosis_Heart_Disease))) +
    geom_boxplot(outlier.color = "red", outlier.shape = 16) +
    labs(title = paste("Boxplot of", var, "by Heart Disease Diagnosis"),
         x = "Heart Diagnosis",
         y = var) +
    theme_minimal()
    print(p)
}
```


Scatterplot

```{r}
pairs(numerical_data)
```

Correlation

```{r}
library(corrplot)

#Using the cor function to find the correlation between the variables
cor_data = cor(numerical_data)
 
print("Correlation matrix")
print(cor_data)
```


```{r}
#correlation matrix 
corrplot(cor_data, method="number", cl.cex = 0.5, tl.col = "blue", tl.cex = 0.5)
```


##### Getting the numerical variables from the data + scaling
```{r}
heart_numerical <- scale(numerical_data)
```


### PCA

```{r}
heart.pca <- prcomp(heart_numerical)

#Eigen vectors or weights

heart.pca
```



```{r}
library(factoextra)
fviz_eig(heart.pca)  ## use the function on the pca object
```



```{r}
library(factoextra)
fviz_pca_biplot(heart.pca, repel = TRUE,
                col.var = "steelblue", # Variables color
                col.ind = "coral")
```

```{r}
#pca loadings
pca_loadings <- heart.pca$rotation
head(pca_loadings)
```

```{r}
#PCA scores
pca_scores <- heart.pca$x
head(pca_scores)

```

PCA prediction 

# Factor Analysis

#### Factor Analysis

#### 1. Check data - Correlations and Sampling Adequacy

###Bartlett Test for correlations

#### H_0: The correlation matrix is the same as the identity matrix
#### H_a: The correlation matrix is different from the identity matrix.

```{r}
head(heart_numerical)
```

```{r}
#run EFA using psych
library(psych)
library(GPArotation)
correl <- cor(heart_numerical)
cortest.bartlett(correl, n = nrow(heart_numerical))
```
#### Since the p-value is less than 0.05, we reject the null hypothesis and conclude that there are significant correlations in the data set.

#### KMO Test for sampling adequacy ( Significant factors in the sampled data)
#### H_0: There are no significant factors in the data set
#### H_a: There are significant factors in the data set.

```{r}
km = KMO(correl)
km$MSA
```

#### Since the overall MSA value is greater than 0.5, we reject the null hypothesis and conclude that there are significant factors in the data set.

#### We have checked the data set and it looks like we can identify some significant factors in the data set. We can now go ahead to perform the factor analysis

#### Load the needed packages
```{r}
library(psych)
library(GPArotation)
```

#### What number of factors do we fit our factor analysis for? We will use the parallel analysis to determine the number of factors. 

#### Extraction method: We will use the maximum likelihood method 
#### Rotations: We will try out all 3 rotations.

#### Number of Factors
```{r}
nofactors <- fa.parallel(numerical_data, fm = "ml", fa = "fa")
```
#### The parallel analysis suggests that we have 3 significant factors in our data set.

```{r}
nofactors$fa.values
```

###Kaiser Criterion
```{r}
nofactors$fa.values[nofactors$fa.values>0.7]
```

#### Factor Analysis

```{r}
df_fa <- fa(heart_numerical, nfactors = 3, fm = "ml", rotate = "none")

df_fa$loadings

fa.diagram(df_fa)
```

```{r}
#### Varimax rotations
df_fa1 <- fa(heart_numerical, nfactors = 3, fm = "ml", rotate = "varimax")

df_fa1$loadings
```

```{r}
fa.diagram(df_fa1)
```

###Oblique
```{r}
df_fa2 <- fa(heart_numerical, nfactors = 3,fm="ml", rotate = "oblimin")

df_fa2
```

```{r}
fa.diagram(df_fa2)
```

# MCA


```{r}
head(categorical_data)
```

```{r}
str(categorical_data)
```


```{r}
# Test for Sex and Diagnosis_Heart_Disease
contingency_sex <- table(categorical_data$Diagnosis_Heart_Disease, categorical_data$Sex)
chi_sex <- chisq.test(contingency_sex)

# Print null and alternative hypotheses for Sex and Diagnosis_Heart_Disease
print("Null Hypothesis (H₀): There is no association between Sex and Diagnosis_Heart_Disease (independent).")
print("Alternative Hypothesis (H₁): There is an association between Sex and Diagnosis_Heart_Disease (dependent).")
print("Chi-square result for Sex and Diagnosis_Heart_Disease:")
print(chi_sex)

# Conclusion based on p-value
if (chi_sex$p.value < 0.05) {
  print("Conclusion: Reject the null hypothesis. There is a significant association between Sex and Diagnosis_Heart_Disease.")
} else {
  print("Conclusion: Fail to reject the null hypothesis. There is no significant association between Sex and Diagnosis_Heart_Disease.")
}

# Test for Exercise_Induced_Angina and Diagnosis_Heart_Disease
contingency_angina <- table(categorical_data$Diagnosis_Heart_Disease, categorical_data$Exercise_Induced_Angina)
chi_angina <- chisq.test(contingency_angina)

# Print null and alternative hypotheses for Exercise_Induced_Angina and Diagnosis_Heart_Disease
print("Null Hypothesis (H₀): There is no association between Exercise_Induced_Angina and Diagnosis_Heart_Disease (independent).")
print("Alternative Hypothesis (H₁): There is an association between Exercise_Induced_Angina and Diagnosis_Heart_Disease (dependent).")
print("Chi-square result for Exercise_Induced_Angina and Diagnosis_Heart_Disease:")
print(chi_angina)

# Conclusion based on p-value
if (chi_angina$p.value < 0.05) {
  print("Conclusion: Reject the null hypothesis. There is a significant association between Exercise_Induced_Angina and Diagnosis_Heart_Disease.")
} else {
  print("Conclusion: Fail to reject the null hypothesis. There is no significant association between Exercise_Induced_Angina and Diagnosis_Heart_Disease.")
}

# Test for Chest_Pain_Type and Diagnosis_Heart_Disease
contingency_chest_pain <- table(categorical_data$Diagnosis_Heart_Disease, categorical_data$Chest_Pain_Type)
chi_chest_pain <- chisq.test(contingency_chest_pain)

# Print null and alternative hypotheses for Chest_Pain_Type and Diagnosis_Heart_Disease
print("Null Hypothesis (H₀): There is no association between Chest_Pain_Type and Diagnosis_Heart_Disease (independent).")
print("Alternative Hypothesis (H₁): There is an association between Chest_Pain_Type and Diagnosis_Heart_Disease (dependent).")
print("Chi-square result for Chest_Pain_Type and Diagnosis_Heart_Disease:")
print(chi_chest_pain)

# Conclusion based on p-value
if (chi_chest_pain$p.value < 0.05) {
  print("Conclusion: Reject the null hypothesis. There is a significant association between Chest_Pain_Type and Diagnosis_Heart_Disease.")
} else {
  print("Conclusion: Fail to reject the null hypothesis. There is no significant association between Chest_Pain_Type and Diagnosis_Heart_Disease.")
}

# Test for Resting_ECG and Diagnosis_Heart_Disease
contingency_ecg <- table(categorical_data$Diagnosis_Heart_Disease, categorical_data$Resting_ECG)
chi_ecg <- chisq.test(contingency_ecg)

# Print null and alternative hypotheses for Resting_ECG and Diagnosis_Heart_Disease
print("Null Hypothesis (H₀): There is no association between Resting_ECG and Diagnosis_Heart_Disease (independent).")
print("Alternative Hypothesis (H₁): There is an association between Resting_ECG and Diagnosis_Heart_Disease (dependent).")
print("Chi-square result for Resting_ECG and Diagnosis_Heart_Disease:")
print(chi_ecg)

# Conclusion based on p-value
if (chi_ecg$p.value < 0.05) {
  print("Conclusion: Reject the null hypothesis. There is a significant association between Resting_ECG and Diagnosis_Heart_Disease.")
} else {
  print("Conclusion: Fail to reject the null hypothesis. There is no significant association between Resting_ECG and Diagnosis_Heart_Disease.")
}

# Test for Num_Major_Vessels_Flouro and Diagnosis_Heart_Disease
contingency_vessels <- table(categorical_data$Diagnosis_Heart_Disease, categorical_data$Num_Major_Vessels_Flouro)
chi_vessels <- chisq.test(contingency_vessels)

# Print null and alternative hypotheses for Num_Major_Vessels_Flouro and Diagnosis_Heart_Disease
print("Null Hypothesis (H₀): There is no association between Num_Major_Vessels_Flouro and Diagnosis_Heart_Disease (independent).")
print("Alternative Hypothesis (H₁): There is an association between Num_Major_Vessels_Flouro and Diagnosis_Heart_Disease (dependent).")
print("Chi-square result for Num_Major_Vessels_Flouro and Diagnosis_Heart_Disease:")
print(chi_vessels)

# Conclusion based on p-value
if (chi_vessels$p.value < 0.05) {
  print("Conclusion: Reject the null hypothesis. There is a significant association between Num_Major_Vessels_Flouro and Diagnosis_Heart_Disease.")
} else {
  print("Conclusion: Fail to reject the null hypothesis. There is no significant association between Num_Major_Vessels_Flouro and Diagnosis_Heart_Disease.")
}

# Test for Thalassemia and Diagnosis_Heart_Disease
contingency_thalassemia <- table(categorical_data$Diagnosis_Heart_Disease, categorical_data$Thalassemia)
chi_thalassemia <- chisq.test(contingency_thalassemia)

# Print null and alternative hypotheses for Thalassemia and Diagnosis_Heart_Disease
print("Null Hypothesis (H₀): There is no association between Thalassemia and Diagnosis_Heart_Disease (independent).")
print("Alternative Hypothesis (H₁): There is an association between Thalassemia and Diagnosis_Heart_Disease (dependent).")
print("Chi-square result for Thalassemia and Diagnosis_Heart_Disease:")
print(chi_thalassemia)

# Conclusion based on p-value
if (chi_thalassemia$p.value < 0.05) {
  print("Conclusion: Reject the null hypothesis. There is a significant association between Thalassemia and Diagnosis_Heart_Disease.")
} else {
  print("Conclusion: Fail to reject the null hypothesis. There is no significant association between Thalassemia and Diagnosis_Heart_Disease.")
}

# Create contingency table for Age Group and Diagnosis_Heart_Disease
contingency_age_group <- table(categorical_data$Diagnosis_Heart_Disease, categorical_data$Age_Group)

# Perform Chi-square test
chi_age_group <- chisq.test(contingency_age_group)

# Print null and alternative hypotheses for Age Group and Diagnosis_Heart_Disease
print("Null Hypothesis (H₀): There is no association between Age Group and Diagnosis_Heart_Disease (independent).")
print("Alternative Hypothesis (H₁): There is an association between Age Group and Diagnosis_Heart_Disease (dependent).")

# Print Chi-square result for Age Group and Diagnosis_Heart_Disease
print("Chi-square result for Age Group and Diagnosis_Heart_Disease:")
print(chi_age_group)

# Conclusion based on p-value
if (chi_age_group$p.value < 0.05) {
  print("Conclusion: Reject the null hypothesis. There is a significant association between Age Group and Diagnosis_Heart_Disease.")
} else {
  print("Conclusion: Fail to reject the null hypothesis. There is no significant association between Age Group and Diagnosis_Heart_Disease.")
}
```


```{r}
library(FactoMineR)

# apply MCA
mca1 = MCA(categorical_data, graph = FALSE)

# table of eigenvalues
mca1$eig
```
```{r}
# column coordinates
mca1$var$coord
```
```{r}
# row coordinates
mca1$ind$coord
```

```{r}
# number of categories per variable
cats = apply(categorical_data, 2, function(x) nlevels(as.factor(x)))
cats
```

```{r}
library(ggplot2)

# Data frames for ggplot
mca1_vars_df <- data.frame(mca1$var$coord, Variable = rep(names(cats), cats))
mca1_obs_df <- data.frame(mca1$ind$coord)

# Plot of variable categories
ggplot(data = mca1_vars_df, aes(x = Dim.1, y = Dim.2, label = rownames(mca1_vars_df))) +
  geom_hline(yintercept = 0, colour = "gray70") + 
  geom_vline(xintercept = 0, colour = "gray70") +
  geom_text(aes(colour = Variable), size = 3, check_overlap = TRUE) +  # Adjusted size and added overlap check
  ggtitle("MCA plot of variables") +
  theme_minimal() +  # Clean background
  theme(
    plot.title = element_text(size = 14, hjust = 0.5),  # Centered title
    axis.title = element_text(size = 12),  # Axis labels
    axis.text = element_text(size = 10)  # Axis text
  ) +
  xlim(c(-1, 1)) +  # Adjust axis limits for better view
  ylim(c(-1, 1))  # Adjust axis limits for better view
```

```{r}
# MCA plot of observations and categories
ggplot(data = mca1_obs_df, aes(x = Dim.1, y = Dim.2)) + geom_hline(yintercept = 0, 
    colour = "gray70") + geom_vline(xintercept = 0, colour = "gray70") + geom_point(colour = "gray50", 
    alpha = 0.7) + geom_density2d(colour = "gray80") + geom_text(data = mca1_vars_df, 
    aes(x = Dim.1, y = Dim.2, label = rownames(mca1_vars_df), colour = Variable)) + 
    ggtitle("MCA plot of variables using R package FactoMineR") + scale_colour_discrete(name = "Variable")
```
```{r}
# default biplot in FactoMineR
plot(mca1)
```
